---
title: "W4_WrangleViz"
author: "R Club"
date: "2024-02-14"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Valentine's Day!

Today, we're going to work with a data set on speed dating together.

First of all, possibly download (or locate) the data and **load** it into R:

```{r}
# load data into RStudio




```

## What can we do with this data set?

Check out the data set (and possibly the codebook), and decide what aspects (a.k.a. columns/variables) of the data would interest you (as a group!).

195 columns are a lot, so maybe only **keep those columns** that are interesting for you.

```{r}
# check out data


```

## Renaming columns

These column names might not be very helpful to remember what the names stand for. If you want to, try to **change the names** of a few of these columns!

```{r}

```

## Tidying the data

Is the data in a tidy format?

What is a tidy format? (Slides)

If not, **reshape** the data. (And if yes, reshape it anyway to practice! :D)

```{r}

```

(You might merge data frames at some point, possibly here, how do you da that?)

## Calculating the average ratings

We might want to calculate the average of the ratings, e.g. attractiveness across time points.

First of all, unfortunately the ratings are on different scales between waves:

> Waves 6-9: Please rate the importance of the following attributes in a potential date on a scale of 1-10 (1=not at all important, 10=extremely important):
>
> Waves 1-5, 10-21: You have 100 points to distribute among the following attributes -- give more points to those attributes that are more important in a potential date, and fewer points to those attributes that are less important in a potential date.  Total points must equal 100.

1.  **Transform** these ratings to scales that are somewhat comparable!

```{r}

```

1.  Now **calculate the average** ratings across time points per speed dater.

```{r}

```

## Filtering the speed daters

Make a new data frame that only consists of female speed daters! Also filter out those where the partner is older than the speed dater.

```{r}

```

## Sort the data frame

Sort the data frame based on the correlation between the interests, from highest to lowest.

```{r}

```

## Visualize!

(Small Intro –\> Visualization Slides!)

Now decide - as a group - what kind of plot to make!

Make this **plot**, layer by layer.

```{r}

```
