---
title: "W4_WrangleViz"
subtitle: |
   | R Coding Club 
   | RTG 2660
author: "Dr. Lea Hildebrandt"
date: 2024/02/14
description: "Getting the raw data in a useful format"
format: 
  revealjs:
    smaller: true
    scrollable: true
    slide-number: true
    theme: simple
    chalkboard: true
editor: visual
from: markdown+emoji
---

# More Data Wrangling & First Visualizations

## Warm-up

Did anyone bring a problem?

```{r} #| echo: false}
library(tidyverse) 
```

```{css}
code.sourceCode {   font-size: 1.4em; }  
div.cell-output-stdout {   font-size: 1.4em; }
```

. . .

If I type this code into the console (and run it), I will get the following error. Why?

```         
library(babynames)  
dat <- summarise(.data = babynames, mean_n = mean(n))  

Error in summarise(.data = babynames, mean_n = mean(n)) :    could not find function "summarise"
```

. . .

Which function(s) would you use to calculate reaction times per subject?

. . .

How do you merge two data sets?

## Valentines Day!

Let's work with a "romantic" data set today! :D

I found this huge data set on speed dating online, plus it's codebook. Let's check it out.

. . .

First, we need to load the data. For this, there are a variety of functions (also for reading in specialized files). Since it is a basic .csv file, we can use \`read_data()\`:

```{r}
#| echo: false

speed_dating <- read_csv()

```
